Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-07 paper/paper.tex

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\Delta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* ∆)
       (for/fold ([cs* (set)] [∆* '()])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons ∆** cs**)
            (values (set-union cs* cs**) (append ∆** ∆*))])))
     (cons (update ∆ \(\msto\)) (set-union cs cs*))]))
\end{alltt}

@@

  (define (step state)
    (match state
      [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
      [(co \(\msto\) k v)
       (match k
         ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
           (yield (ans \(\msto\) v))))]
         [(ar\(\superscript\mlab\) e \(\menv\))
          (generator (do (\(\msto\)) ()
            (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
         [(fn\(\superscript\mlab\) f)
          (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
           (yield (ap \(\msto\) f v k))))]
         [(fi\(\superscript\mlab\) c a \(\menv\))
          (generator
           (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                    [v <- (force \(\msto\) v)])
             (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
      [(ap \(\msto\) fun a k)
       (match fun
         [(clos l x e \(\menv\))
          (generator
           (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
             (yield (ev \(\msto\)* e \(\menv\)* k))))]
         [(? op? o)
          (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                   [v <- (force \(\msto\) v)]
                   [v (\(\Delta\) o (list v))])
            (yield (co \(\msto\) k v)))
         [_ (generator (do (\(\msto\)) () (continue)))]))]))
\end{alltt}
\caption{Parameterized Abstract$^2$ machine for ISWIM}
\label{fig:paam}
\end{figure}

@@

Why generators? Relations in general cannot be written as computable
functions, but in our case we have a decidable, finitely
non-deterministic relation. There are several implementation
strategies for relations, some more effective than others. For
instance, a silly implementation strategy would be defining a binary
function that decides membership in the relation. We want the
definition to drive the interpretor, so we might naturally decide to
write a function that takes a state and returns a set of all the next
states. As we described in the imperative workset section, this
intermediate allocation is unnecessary and wasteful. What we really
want is a way of pulling on the reduction relation for more next
states as the fixpoint is ready for them.

@@

There are a few steps to consider:
\begin{itemize}
 \item{Define special value lattice elements for compound data domains that can be quoted
       (e.g. {\tt Pair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt IVector} for immutable vectors, etc.)}
 \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
 \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
       and similar definitions for immutable vectors.}
 \item{Extend the $\Delta$ axioms to include conservative meaning for these new values (e.g. {\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
\end{itemize}

\begin{figure}
\begin{align*}
\Delta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
\Delta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \lbrace \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\rbrace] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
merge(v, v) &= v \\
merge(n, m) &= {\tt Number} \\
merge(n, v) &= {\tt QData} \\
merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt Pair} \\
merge({\tt Pair}, ({\tt cons}\ a\ d)) &= {\tt Pair} \\
merge({\tt Pair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}

>>>>

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* ∆)
       (for/fold ([cs* (set)] [∆* '()])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons \(\msdiff\)** cs**)
            (values (set-union cs* cs**) (append \(\msdiff\)** \(\msdiff\)*))])))
     (cons (update \(\msdiff\) \(\msto\)) (set-union cs cs*))]))
\end{alltt}

@@

(define (step state)
  (match state
    [(ev \(\msto\) e \(\menv\) k) \(e-dispatch\)] ;; dead if compiled.
    [(co \(\msto\) k v)
     (match k
       ['mt (generator (do (\(\msto\)) ([v <- (force \(\msto\) v)])
        (yield (ans \(\msto\) v))))]
       [(ar\(\superscript\mlab\) e \(\menv\))
        (generator (do (\(\msto\)) ()
          (yield (ev \(\msto\) e \(\menv\) (fn v l)))))
       [(fn\(\superscript\mlab\) f)
        (generator (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)])
         (yield (ap \(\msto\) f v k))))]
       [(fi\(\superscript\mlab\) c a \(\menv\))
        (generator
         (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                  [v <- (force \(\msto\) v)])
           (yield (ev \(\msto\) (if v c a) \(\menv\) k))))])]
    [(ap \(\msto\) fun a k)
     (match fun
       [(clos l x e \(\menv\))
        (generator
         (do (\(\msto\)) ([(\(\menv\)* \(\msto\)*) (bind state)])
           (yield (ev \(\msto\)* e \(\menv\)* k))))]
       [(? op? o)
        (do (\(\msto\)) ([k <- (get-cont \(\msto\) l)]
                 [v <- (force \(\msto\) v)]
                 [v (\(\interpdelta\) o (list v))])
          (yield (co \(\msto\) k v)))
       [_ (generator (do (\(\msto\)) () (continue)))]))]))
\end{alltt}
\caption{Parameterized Abstract$^2$ machine for ISWIM}
\label{fig:paam}
\end{figure}

@@

Why generators? Relations in general cannot be written as computable
functions, but in our case we have a decidable relation with finitely
many pairs sharing the same left component. There are several
implementation strategies for such relations, some more effective than
others. For instance, a silly implementation strategy would be
defining a binary function that decides membership in the relation. We
want the definition to drive the interpreter, so we might naturally
decide to write a function that takes a state and returns a set of all
the next states. As we described in the imperative workset section,
this intermediate allocation is unnecessary and wasteful. What we
really want is a way of pulling on the reduction relation for more
next states as the fixpoint is ready for them.

@@

There are a few steps to consider:
\begin{itemize}
 \item{Define special value lattice elements for compound data domains that can be quoted
       (e.g. {\tt QPair} for $\left\lbrace ({\tt cons} a_i d_i)\right\rbrace_i$, {\tt QVector} for immutable vectors, etc.)}
 \item{Define a ``larger'' value lattice element for all quotable data, {\tt QData}}
 \item{Interpret {\tt (quote (a ...))} as {\tt (qlist a ...)}, a new primitive function defined in figure \ref{fig:qlist},
       and similar definitions for immutable vectors.}
 \item{Extend the $\interpdelta$ axioms to include conservative meaning for these new values (e.g. {\tt (car QData)} = {\tt QData}, {\tt (add1 QData)} = {\tt Number} and log ``possible type error'') and allow them to allocate addresses and change the store}
\end{itemize}

\begin{figure}
\begin{align*}
\interpdelta(\msto, qlist) &= (\mbox{{\tt '()}}, \msto) \\
\interpdelta(\msto, qlist, v ..._+) &= (({\tt cons}\ a\ d), \msto') \\
\qquad \mbox{where } \msto' &= \msto\sqcup[a \mapsto \bigsqcup(v ...)] \\
                            &\qquad \sqcup[d \mapsto \lbrace \mbox{{\tt '()}}, ({\tt cons}\ a\ d)\rbrace] \\
\bigsqcup(v) &= v \\
\bigsqcup(v, vs ...) &= merge(v, \bigsqcup(vs ...)) \\
merge(v, v) &= v \\
merge(n, m) &= {\tt Number} \\
merge(n, v) &= {\tt QData} \\
merge(({\tt cons}\ a\ d), ({\tt cons}\ a'\ d')) &= {\tt QPair} \\
merge({\tt QPair}, ({\tt cons}\ a\ d)) &= {\tt QPair} \\
merge({\tt QPair}, v) &= {\tt QData} \\
\vdots
\end{align*}
\caption{Quoted list primitive}
\label{fig:qlist}
\end{figure}
